---
title: "Cheatsheet of the tutorials"
author: "Martin Raden"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

This page aggregates the cheat sheets of our tutorials and additional information from out
R tutorial package [Introduction to programming with R](https://github.com/Dr-Eberle-Zentrum/Introduction-to-programming-with-R).


## R, RStudio, coding, packages

- `==` operator tests for equality of things; `!=` for inequality
- `=` is used for assignment of values to arguments (don't use for object assignment!)
- Boolean statements work on and result in `TRUE/FALSE` values 
  - can be combined using logical operators like `&` (AND), `|` (OR), and `!` (NOT)
- *functions*, aka *commands*, are the workhorses in R and always come with a `()` after their name
- `c()` combines/concatenates multiple values into one *vector* object
- `install.packages()` installs a new package via its name from the console (also possible via 'Packages' menu in RStudio)
- a package is a collection of additional functions and data objects, examples are
  - `ggplot2` for visualization
  - `dplyr` for data transformation
  - important packages for data processing covered in this course are loaded via the meta-package `tidyverse`
- `library()` loads a package via its name
  - this command is *typically missing* if your R doesn't know/find the function you want to use!



```{r child = 'inst/tutorials/first-steps/images/cheatsheet.Rmd'}
```

```{r child = 'inst/tutorials/ggplot-basics/images/cheatsheet.Rmd'}
```

## Workflow basics

- variable names 
  - can contain numbers, letters, `.` and `_`
  - have to start with a letter
  - should be self-explanatory
  - capitalization matters, i.e. `myname` is not `myName`
- use `<-` for object assignment (RStudio hot key = 'ALT' + '-')
- `seq()` generates a vector of subsequent numbers, e.g. `seq(2,5)` is equal to `c(2,3,4,5)`
- text/string/character values have to be *quoted*, i.e. surrounded by ticks
  - use either `''` or `""`, both works
  - ensure closing tick for any opening
- writing a variable name in the console prints it, i.e. `x` is equal to `print(x)`
  - surrounding a term in round brackets `()` triggers printing implicitly
- RStudio: hot key 'ALT'+'SHIFT'+'K' shows all hot key combinations


```{r child = 'inst/tutorials/dplyr-filter/images/cheatsheet.Rmd'}
```

```{r child = 'inst/tutorials/dplyr-arrange-select/images/cheatsheet.Rmd'}
```

```{r child = 'inst/tutorials/dplyr-mutate/images/cheatsheet.Rmd'}
```

```{r child = 'inst/tutorials/dplyr-groupby-summarize/images/cheatsheet.Rmd'}
```


--------------------------------------------------------


This work is licensed under a [Creative Commons Attribution-ShareAlike 4.0 International License](http://creativecommons.org/licenses/by-sa/4.0/).

Feel free to contribute at [https://github.com/Dr-Eberle-Zentrum/Introduction-to-programming-with-R](https://github.com/Dr-Eberle-Zentrum/Introduction-to-programming-with-R).

