---
title: "Algorithmic thinking"
tutorial:
  id: "de.uni-tuebingen.dezfdk.tutorials.r-algorithmic-thinking"
  version: 1.0
  author: Martin Raden
output: 
  learnr::tutorial:
    progressive: false
    allow_skip: true
    theme: "sandstone"
    highlight: haddock
    language: "en"
    includes:
      in_header: "../dez-style.html"
      before_body: "../logo-header.html"
      after_body: "../footer.html"
runtime: shiny_prerendered
---
  
```{r setup, include=FALSE}
library(learnr)
```

```{css echo=FALSE}
```

```{js print2pdf1, context="server", echo=FALSE}
// the following 2 chunks print the completed sections of the tutorial to PDF
// uses "css/print2pdf.css"
$(document).on('shiny:inputchanged', function(event) {
  if (event.name === 'print2pdf') {
    window.print();
  }
});
```

## How to start?

Most of the major initial questions of programming novices are *Where do I start? How to begin? How to tackle the problem?*

Thus, before we start to get deeper into any programming or data processing, we first have to train our mind set for *algorithmic problem solving*. This is also known as *algorithmic* or *computational thinking*, which is briefly introduced in the following [video](https://youtu.be/_TWsmF8l418).

![](https://youtu.be/_TWsmF8l418)


https://therenegadecoder.com/code/algorithmic-thinking-for-python-beginners/

## General remarks

When given a task to be solved via programming, the following general remarks might be helpful!

- FIRST, make A PLAN! That is, you have to make written notes to identify individual MAJOR steps/parts. 
- Once you have the first plan: RESTART DECOMPOSITION, i.e. revise your steps whether they can be SPLIT INTO DETAILED steps. 

## Print

You can print (e.g. to PDF file) your current processing state of this tutorial, i.e. all the already visited topics, using the button from below.

```{r print2pdf2, echo=FALSE}
# button can be placed anywhere in the tutorial
actionButton("print2pdf", "Print processed tutorial parts", style="opacity: .7; color: #000;")
```

To this end, you have to open the tutorial within a standard web browser like Firefox, Chrome, Safari, Edge, ... When you are running the tutorial within RStudio, you can use the `Open in Browser` button in the upper left corner of the window that shows the tutorial.
